---
title: "id_verification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{id_verification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ohcleandat)
library(dplyr)
```

## The problem

Before two datasets can be integrated, the IDs that link them have to match.
Some mismatches are expected in field-collected data: a household recorded in
a sampling roster but not yet in the survey, a typo introduced during entry, a record
entered twice.

`create_id_log()` takes two "semi-clean" datasets and returns a **validation
log** — one row per problem ID, formatted so a human can review it, decide what
the correct value is, and initial the change.

## Example data

Two small datasets linked by household ID. The household file uses `hh_id`;
the individual roster calls the same thing `household_id`.

```{r}
households <- tibble::tibble(
  hh_record = 1:4,
  hh_id     = c("H001", "H002", "H003", "H004"),
  village   = c("Kisumu", "Kisumu", "Siaya", "Siaya")
)

sampling_roster <- tibble::tibble(
  person_record = 1:5,
  household_id  = c("H001", "H002", "H002", "H005", "H006"),
  age           = c(34, 12, 41, 27, 8)
)
```

`H003` and `H004` have no one in the sampling roster. `H005` and `H006` are in the
sampling roster but not in the household file. Those four are what we want flagged.

## Creating the log

```{r creating log}
# id_log <- ohcleandat::create_id_log(
#   semiclean_x   = households,
#   semiclean_y   = roster,
#   by            = c("hh_id" = "household_id"),
#   primary_key_x = "hh_record",
#   primary_key_y = "person_record",
#   name_x        = "data/raw/households_2026.csv",
#   name_y        = "data/raw/roster_2026.csv"
# )
# 
# id_log
```

Each row is one unmatched ID. The columns are:

| Column | What it holds |
|---|---|
| `dataset` | Which file the problem ID came from |
| `entry` | The primary key of the offending row, so you can find it |
| `field` | The ID column being checked |
| `issue` | A plain-language description of the mismatch |
| `old_value` | The ID as it currently appears |
| `no_change`, `new_val`, `user_initials`, `comments` | Blank — filled in by the reviewer |

The last four are intentionally empty. The log is meant to be written out,
handed to whoever knows the data, and returned with decisions recorded.

Although `create_id_log()` compares two datasets at a time, a study rarely stops at
two. When the same ID appears across several files — a household ID shared by
the survey, the roster, and the visit records — run the function once per pair
and combine the results with `dplyr::bind_rows()`. The output schema is the
same every time, and the `dataset` column identifies where each flagged ID came
from, so a combined log reads as one review task rather than several. Which
pairs to run is a study design question, not a technical one: if one file serves
as the authoritative source of IDs, compare each of the others against it and
leave it at that; if the files are meant to be integrated in sequence, compare
each pair in the order they will be joined. Comparing every possible pair is an
option, but on more than three or four datasets it produces overlapping rows
that ask a reviewer to resolve the same ID more than once. Match the set of
comparisons to the integration you actually intend to perform.

## Handing it off for review

```{r handing off, eval = FALSE}
readr::write_csv(id_log, "logs/id_validation_2026-08-19.csv")
```

A reviewer marks `no_change` for IDs that are legitimately absent (a household
that genuinely enrolled no one) with a "T" or "True", or enters a "F" or "False"
 and fills `new_val` with the corrected ID where it's a typo. 
`user_initials` and `comments` keep the audit trail.

## Matching columns with the same name

When both datasets use the same column name, `by` is just a string:

```{r, eval = FALSE}
create_id_log(
  semiclean_x   = households,
  semiclean_y   = roster,
  by            = "hh_id",
  primary_key_x = "hh_record",
  primary_key_y = "person_record",
  name_x        = "households.csv",
  name_y        = "roster.csv"
)
```

## Where this fits

`create_id_log()` is a another checking step that gets integrating into the ohcleandat
cleaning mechanics. The intended loop is: run the log, review it, apply the corrections to
your semi-clean files, then run it again. When the all the rows within the log 
have been filled out, the cleaning task is compelte.

## See also

`dplyr::anti_join()`, which does the underlying comparison.
