---
title: "id_verification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{id_verification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(ohcleandat)
library(dplyr)
```

## The problem

Before two datasets can be integrated, the IDs that link them have to match.
Some mismatches are expected in field-collected data: a household recorded in
a sampling roster but not yet in the survey, a typo introduced during entry, a record
entered twice.

`create_id_log()` takes two "semi-clean" datasets and returns a **validation
log** — one row per problem ID, formatted so a human can review it, decide what
the correct value is, and initial the change. The function checks for two
classes of problem: IDs that exist in one dataset but not the other, and IDs
that appear more than once within a single dataset.

## Example data

Two small datasets linked by household ID. The household file uses `hh_id`;
the individual roster calls the same thing `household_id`.

```{r}
households <- tibble::tibble(
  hh_record = 1:5,
  hh_id     = c("H001", "H002", "H003", "H004", "H002"),
  village   = c("Kisumu", "Kisumu", "Siaya", "Siaya", "Kisumu")
)

sampling_roster <- tibble::tibble(
  person_record = 1:5,
  household_id  = c("H001", "H002", "H002", "H005", "H006"),
  age           = c(34, 12, 41, 27, 8)
)
```

`H003` and `H004` have no one in the sampling roster. `H005` and `H006` are in
the sampling roster but not in the household file. Those four are what we want
flagged as cross-dataset mismatches. In addition, `H002` appears twice in the
household file — that duplicate is flagged separately.

## Creating the log

```{r creating log, eval=FALSE}
id_log <- ohcleandat::create_id_log(
  semiclean_x   = households,
  semiclean_y   = roster,
  by            = c("hh_id" = "household_id"),
  primary_key_x = "hh_record",
  primary_key_y = "person_record",
  name_x        = "data/raw/households_2026.csv",
  name_y        = "data/raw/roster_2026.csv"
)

id_log
```

Each row is one problem ID. The columns are:

| Column | What it holds |
|---|---|
| `dataset` | Which file the problem ID came from |
| `entry` | The primary key of the offending row, so you can find it |
| `field` | The ID column being checked |
| `issue` | A plain-language description of the mismatch or duplication |
| `old_value` | The ID as it currently appears |
| `no_change`, `new_val`, `user_initials`, `comments` | Blank — filled in by the reviewer |

The last four are intentionally empty. The log is meant to be written out,
handed to whoever knows the data, and returned with decisions recorded.

## Duplicate IDs

In addition to cross-dataset mismatches, `create_id_log()` checks each dataset
independently for IDs that appear more than once. Duplicates are flagged in the
same log under a distinct issue description, so a reviewer sees all ID problems
in one place.

A duplicate row in the log does not mean the record should be deleted. It means
the same ID was recorded more than once and a human needs to decide whether the
two rows represent the same event entered twice, or two genuinely different
records that were assigned the same ID by mistake. The reviewer uses `no_change`
and `new_val` to record that decision, just as they would for a cross-dataset
mismatch.

```{r duplicates, eval=FALSE}
# H002 appears twice in households — both rows will appear in the log
# with issue = "household_id in dataset_x does not match any HHID in dataset_y"
id_log |>
  dplyr::filter(grepl("does not match", issue)) 
```

When the same ID is duplicated in both datasets, two separate rows appear in
the log — one flagged against each dataset — so the reviewer can assess each
side independently.

## Handing it off for review

```{r handing off, eval = FALSE}
readr::write_csv(id_log, "logs/id_validation_2026-08-19.csv")
```

A reviewer marks `no_change` for IDs that are legitimately absent (a household
that genuinely enrolled no one) with a "T" or "True", or enters a "F" or "False"
and fills `new_val` with the corrected ID where it's a typo.
`user_initials` and `comments` keep the audit trail. The same convention applies
to duplicates: if both rows are legitimate, the reviewer assigns a corrected ID
to one of them in `new_val`; if one is a true duplicate entry, they mark it
`no_change = "T"` to indicate it can be dropped.

## Matching columns with the same name

When both datasets use the same column name, `by` is just a string:

```{r, eval = FALSE}
create_id_log(
  semiclean_x   = households,
  semiclean_y   = roster,
  by            = "hh_id",
  primary_key_x = "hh_record",
  primary_key_y = "person_record",
  name_x        = "households.csv",
  name_y        = "roster.csv"
)

# by is a character variable containing match id, or if named different, 
# a named character vector like c("a" = "b")
# this function assumes that the named character vector follows the convention of
# c("variable in semiclean_x" = "variable in semiclean_y")
```

## Combining logs across multiple datasets

Although `create_id_log()` compares two datasets at a time, a study rarely stops
at two. When the same ID appears across several files — a household ID shared by
the survey, the roster, and the visit records — run the function once per pair
and combine the results with `dplyr::bind_rows()`. The output schema is the
same every time, and the `dataset` column identifies where each flagged ID came
from, so a combined log reads as one review task rather than several. Which
pairs to run is a study design question, not a technical one: if one file serves
as the authoritative source of IDs, compare each of the others against it and
leave it at that; if the files are meant to be integrated in sequence, compare
each pair in the order they will be joined. Comparing every possible pair is an
option, but on more than three or four datasets it produces overlapping rows
that ask a reviewer to resolve the same ID more than once. Match the set of
comparisons to the integration you actually intend to perform.

## Where this fits

`create_id_log()` is another checking step that gets integrated into the ohcleandat
cleaning mechanics. The intended loop is: run the log, review it, apply the
corrections to your semi-clean files, then run it again. When all the rows
within the log have been filled out, the cleaning task is complete.

## See also

`dplyr::anti_join()`, which does the underlying cross-dataset comparison.
